Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-08-27 analysis/paper/paper.Rmd
fixed equations

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification biases
(Supplement B) and (optionally) a delayed reproduction term to account
for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum
potential recruitment of female calves per cow $\dot{X}_t$:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation} where $w$ is the ratio of observed adult females to
collared animals, $\dot{q}$ is the ratio of young bulls to adult
females, $\dot{u}$ is the probability of misidentifying young bulls as
adult females and vice versa, and $\dot{z}$ is the probability of
missing a calf. We examine sensitivity to these unknown parameters by
selecting values for each simulated example population from plausible
ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and
$\dot{q}$ varies uniformly between 0 and 0.6.

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females
and the adjusted recruitment rate: \begin{equation}
  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
  (\#eq:dotJ)
\end{equation}

Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
population sizes, and declines to $0.6\dot{X}_t$ as the population
reaches carrying capacity of $k=100$ times the initial population size.
The post-juvenile female population in the next year includes both
survivors and new recruits: \begin{equation}
\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
  (\#eq:dotN)
\end{equation}

@@

```{=tex}
\begin{equation}
\bar{R}_t \sim Beta(\mu^R_t,\phi^R); 
log(\mu^R_t) = \dot{\beta}^R_0+\dot{\beta}^R_a A_t+\dot{\beta}^R_f F_t,
  (\#eq:barR)
\end{equation}
```
```{=tex}
\begin{equation}
\bar{S}_t \sim (46 \times Beta(\mu^S_t,\phi^S)-0.5)/45;
log(\mu^S_t) = \dot{\beta}^S_0+\dot{\beta}^S_a A_t.
  (\#eq:barS)
\end{equation}
```
$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period.

>>>>

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification biases
(Supplement B) and (optionally) a delayed reproduction term to account
for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum
potential recruitment of female calves per cow $\dot{X}_t$:

\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}

where $w$ is the ratio of observed adult females to
collared animals, $\dot{q}$ is the ratio of young bulls to adult
females, $\dot{u}$ is the probability of misidentifying young bulls as
adult females and vice versa, and $\dot{z}$ is the probability of
missing a calf. We examine sensitivity to these unknown parameters by
selecting values for each simulated example population from plausible
ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and
$\dot{q}$ varies uniformly between 0 and 0.6.

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females
and the adjusted recruitment rate:

\begin{equation}
  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
  (\#eq:dotJ)
\end{equation}

Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
population sizes, and declines to $0.6\dot{X}_t$ as the population
reaches carrying capacity of $k=100$ times the initial population size.
The post-juvenile female population in the next year includes both
survivors and new recruits:

\begin{equation}
\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
  (\#eq:dotN)
\end{equation}

@@

\begin{equation}
\bar{R}_t \sim Beta(\mu^R_t,\phi^R); 
log(\mu^R_t) = \dot{\beta}^R_0+\dot{\beta}^R_a A_t+\dot{\beta}^R_f F_t,
  (\#eq:barR)
\end{equation}

\begin{equation}
\bar{S}_t \sim (46 \times Beta(\mu^S_t,\phi^S)-0.5)/45;
log(\mu^S_t) = \dot{\beta}^S_0+\dot{\beta}^S_a A_t.
  (\#eq:barS)
\end{equation}

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period.
